Reasonably Programmable Literal Notation

### 2018-02-25 icfp18/syntax-icfp18.tex
first four pages are in decent shape and page breaks are in pleasant locations _twitch_

@@

\vspace{-5px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures. For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expressions and patterns \cite{harper1997programming,mthm97-for-dart}, and many other general-purpose languages support substantially similar list literal notation. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15}. The example in Fig. \ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-4px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-8px}
\end{figure}

This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps, hash tables and so on, following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of string injection attacks, which are amongst the most common and serious security vulnerabilities in web applications \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar. Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than strings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.

\paragraph{Problem} The problem that motivates this paper is that existing syntax extension systems make it difficult or impossible to reason abstractly, i.e. by inspecting only the program as presented to the programmer, about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the implementation details of the extensions that may have been responsible for producing that expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[0.96\textwidth][l]{\textsc{existing grammar extension systems}}}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
  EXTEND /* loaded by, e.g., camlp4 */
    expr:
      |  "`(" q = kquery ")`" -> q
    kquery: 
      | /* ... kquery grammar ... */
  /* ...more extensions defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t 
    lexer  KQueryLexer
    parser KQueryParser.start
    expansions require
      module KQuery as KQuery;
  /* ...more notations defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and making the type and binding structure and the segmentation explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
\vspace{3px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} -- Lines 1-5 use the grammar extension syntax from Camlp4, but see SugarHaskell for another design \cite{erdweg2012layout}. 
The literal form on Line 9 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot abstractly answer questions like the following:

\begin{enumerate}[nolistsep,noitemsep,leftmargin=12px]
\item[\Circle] \textbf{Responsibility}: Which syntax extension determined the expansion of the literal form on Line 9? Might activating a new extension generate a conflicting expansion for the same form?
\item[\Circle] \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item[\Circle] \textbf{Capture}: If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
\item[\Circle] \textbf{Context Dependence}: If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item[\Circle] \textbf{Typing}: What type does \li{y} have? What type is each spliced expression expected to have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. Ideally, the programmer would be able to abstract in some analagous manner over the implementation of a new literal notation. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against. So with TSLs, Line 9 of our example would require a type annotation:

@@

  let y : KQuery.t = `(!R)@&{&/x!/:2_!x}'!R`

@@

The parser associated with the type \li{KQuery.t} would be statically invoked to parse and expand the literal body, i.e. the characters between the outer delimiters, here \li{`(} and \li{)`}, which are left unparsed by the fixed context-free syntax of the language (i.e. generalized literal forms are initially parsed like raw string literals.) TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.

@@

  \item[\CIRCLE] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form -- the parser defined when the type was defined. The context-free syntax of the language remains fixed and composition is mediated by splicing, so clients do not need to worry about different TSLs conflicting syntactically with one another. 
  \item[\Circle] \textbf{Segmentation}: The TSL can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e. subsequence) of the literal body, nor is there a guarantee that spliced expressions are non-overlapping.
  \item[\Circle] \textbf{Capture}: The TSL mechanism as described does not guarantee that splicing is capture-avoiding. (We will return to say more about this problem later.)

@@

  \item[\LEFTcircle] \textbf{Typing}: Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible. However, there is no easy way to determine the type expected for each spliced expression. In addition, the prior work on TSLs considered only a monomorphic, nominally-typed language with only local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml \cite{ocaml-manual}:
\begin{enumerate}[leftmargin=15px]
  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types.
  \item \textbf{Parameterized Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. at all types \li{list('a)}.
  \item \textbf{Abstract Types}: There is no way to define literal notation over a module-parameterized family of abstract types, e.g. at every abstract type defined by a module implementing the \li{QUEUE} signature. Parameterized and abstract types are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
  \item \textbf{ML-Style Type Inference}: The type-directed dispatch mechanism is specified by a bidirectional type system, which supports only local type inference \cite{Pierce:2000:LTI:345099.345100}. It is not clear that this is suitable for languages with ML-style constraint-based type inference.
  \item \textbf{Pattern Literals}: TSLs support only expression literals. No consideration was given to the problem of defining new literal notation in pattern position. Pattern matching is ubiquitous in ML-family languages. %Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.

@@

We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}. All together, we found that no existing approach was suitable as-is for integration into Reason/OCaml.

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): a new system for defining literal notation that addresses all of these deficiencies. The result is the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages. We evaluate the first claim by developing a series of formal calculi equipped with proofs of these abstract reasoning principles, and the second claim with a number of non-trivial examples that appear throughout the paper and involve the language features marked {(a)-(e)} above. In describing these examples, we also demonstrate that the literal parsers can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems for integrating new literal notation into a programming language.

@@

The most obvious point of departure between TSLs, described above, and TLMs, which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are instead explicitly named and applied. For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example just discussed. The programmer can reason abstractly as follows:
\begin{enumerate}[leftmargin=12pt]
\item[\CIRCLE] \textbf{(Responsibility)} The parser named by applied TLM, here \li{KQueryParser} (Line 1), is exclusively responsible for parsing the literal body and generates an expansion statically. Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type. As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, shown in a later example.)
\item[\CIRCLE] \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item[\CIRCLE] \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item[\CIRCLE] \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}. The external dependencies of the generated expansion are explicit -- here, the expansion uses the module \li{Helpers} -- and the system ensures that this dependency is properly bound even if the identifier \li{Helpers} has been shadowed at the application site.
\item[\CIRCLE] \textbf{(Typing)} The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown). Moreover, each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference can be performed abstractly, i.e. given only the inferred segmentation rather than the full expansion\todo{say something formally about this?}. This segment-specific type information can also be communicated to the programmer upon request via the Merlin editor service. We will see examples of TLMs that make use of more advanced features of the OCaml type system in subsequent examples. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\end{enumerate}

>>>>

\vspace{-5px}

@@

\vspace{-5px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
\vspace{-2px}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures. For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}. Many other languages have similar list literal notation. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}. The example in Fig. \ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

\begin{figure}[h]
\vspace{-11px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-5px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-9px}
\end{figure}

This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history,TSLs}. For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps, hash tables and so on, following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of catastrophic string injection attacks~\cite{owasp2017}. More generally, evidence suggests that programmers frequently resort to ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar.
%, e.g. for URLs, paths, regular expressions and so on
Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of this data, which are more amenable to programmatic manipulation and compositional reasoning. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and other systems that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms or type declaration forms, which are beyond the scope of this paper).

\paragraph{Problem} The problem that specifically motivates this paper is that existing syntax extension systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the implementation details of the extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
\vspace{-12px}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
  EXTEND /* loaded by, e.g., camlp4 */
    expr:
      |  "`(" q = kquery ")`" -> q
    kquery: 
      | /* ...K query grammar... */
  /* ...more extensions defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
\vspace{-12px}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t 
    lexer  KQueryLexer
    parser KQueryParser.start
    expansions require
      module KQuery as KQuery;
  /* ...more notations defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and making the type and binding structure and the segmentation explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
% \vspace{4px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system -- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}. 
The literal form on Line 9 then constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e. holding abstract) the details elided on Lines 5-6, cannot easily answer questions like the following:

\begin{enumerate}[leftmargin=12px]
\item[\Circle] \textbf{Responsibility}: Which syntax extension even determines the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same form?
\item[\Circle] \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line 9 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item[\Circle] \textbf{Capture}: If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 9?
\item[\Circle] \textbf{Context Dependence}: If \li{w}, on Lines 7-8, is renamed, could that break Line 9  because the expansion assumes that \li{w} is in scope? What other hidden assumptions is the expansion making?
\item[\Circle] \textbf{Typing}: What type does \li{y} have? What type is each spliced expression expected to have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. Ideally, the programmer would be able to abstract in some analagous manner over the implementation of an unfamiliar notation. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

An approach that has made some headway toward this ideal is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. A language that supports TSLs delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parser associated with the type that the form is being checked against. So in a language with TSLs, our example requires a type annotation on \li{y}:

@@

  let y : KQuery.t = `((!R)@&{&/x!/:2_!x}'!R)`

@@

The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e. the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}. The literal body is constrained by the context-free syntax of the language only in that nested delimiters must be matched (like comments in OCaml). TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.

@@

  \item[\LEFTcircle] \textbf{Responsibility}: The type-directed dispatch mechanism allows the client to easily determine which parser is responsible for each literal form: the parser that was associated with the type when it was defined. Clients do not need to worry about different TSLs conflicting syntactically because the context-free syntax of the language remains fixed and composition is mediated by splicing. The main limitation here is that it is impossible to define literal notation after a type has been defined, and it is also impossible to define multiple notations at a single type.
  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e. subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.
  \item[\LEFTcircle] \textbf{Capture}: The TSL mechanism as described enforces complete capture avoidance, though the formal specification given in the paper did not adequately enforce this constraint.

@@

  \item[\LEFTcircle] \textbf{Typing}: Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible. However, there is no easy way to determine the type expected for each spliced expression. In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml's semantics \cite{ocaml-manual}:
\begin{enumerate}[leftmargin=15px,nolistsep,noitemsep]
  \item \textbf{Structural Types}: There is no way to define literal notation at structural types, e.g. tuple and arrow types, because there is no ``definition site'' for such types.
  \item \textbf{Parameterized \& Abstract Types}: There is no way to define literal notation over a type-parameterized family of types, e.g. at all types \li{list('a)}. Similarly, there is no way to define literal notation over a module-parameterized family of abstract types, e.g. at every abstract type defined by a module implementing the \li{QUEUE} signature. Parameterized and abstract type families are ubiquitous in ML-family languages.% Programmers in ML-family languages commonly define various implementations of abstract data types in this way.
  \item \textbf{Pattern Literals}: Pattern matching is ubiquitous in ML-family languages but the prior work on TSLs considered only expression literals. %Pattern literals are dual to expression literals in that expression literals support value construction, while pattern literals support value deconstruction.
  \item \textbf{ML-Style Type Inference}: It is not clear that a type-directed dispatch scheme can be cleanly  reconciled with ML-style type inference.

@@

We consider other existing approaches against these criteria in Sec.~\ref{sec:existing-approaches}. Taken together, we found that no existing approach was reasonable and expressive enough for integration into Reason/OCaml.

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages. We evaluate these claims with (1) a series of increasingly ML-like calculi that are equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above. In describing these examples, we also demonstrate that literal parsing logic can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems. % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable. 
Unique to our approach in this paper is the focus, both formal and informal, on maintaining client-side abstract reasoning principles.

@@

The most obvious point of departure between TSLs, described above, and TLMs, which are the topic of this paper, is that we separate out the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are instead explicitly named and applied. For example, Line 9 of Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kq} to a generalized literal form delimited by \li{`(} and \li{)`} to express the K query example. The programmer can reason abstractly as follows:
\begin{enumerate}[leftmargin=12pt]
\item[\CIRCLE] \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 is exclusively responsible for lexing, parsing and expanding the literal body. We will give more details on constructing a TLM lexer and parser in the next section. Due to this explicit dispatch mechanism, there can be any number of TLMs defined at a given type. As with TSLs, the context-free syntax of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item[\CIRCLE] \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation is all that needs to be communicated to editor services downstream of the expander, and ultimately to the programmer using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are green).
\item[\CIRCLE] \textbf{Capture}: Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item[\CIRCLE] \textbf{Context Dependence}: The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}. The external dependencies of the generated expansion are explicit -- here, Lines 4-5 specify that the expansion requires the module \li{KQuery} -- and the system ensures that this dependency is properly bound even if the identifier \li{KQuery} has been shadowed at the application site.
\item[\CIRCLE] \textbf{Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion is validated against, here \li{KQuery.t} (not shown). Moreover, each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. given only the inferred segmentation rather than the full expansion. We will see examples of TLMs that make use of more advanced features of the OCaml type system -- notably, pattern matching, parameterized types and modules -- later. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\end{enumerate}
